---
title: "Subscribing Customer Classification"
author: "Olufunke Ojosipe"
date: '2023-03-29'
output:
  pdf_document: default
  html_document: default
---

## Introduction.
You have been provided with marketing data from a bank that has been making phone calls to customers asking them to subscribe a product service.

The object of this notebook is to create a system to predict whether an individual will subscribe to the new line of product. The bank will in turn use this system to target individuals who are more likely to subscribe to the new line of product.

**Load required libraries**
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate)
library(ggplot2) 
library(corrplot)
library(gridExtra)

library(caret)
library(randomForest)
library(party)

require(tree)
library(vip)
```

## Exploratory Data Analysis
This section of the notebook explores the data to find insights, patterns and correlations that can help make informed decision in the model selection and development.

**Check and set working directory**
```{r set directory}
knitr::opts_chunk$set(echo = TRUE)

getwd()
setwd('C:/Users/path-to-folder')
```

**import data**
```{r import data}
knitr::opts_chunk$set(echo = TRUE)

df <- read.csv(file ='train.csv')
```

### Basic Checks
**Perform basic checks on the data**
```{r glimpse data}
knitr::opts_chunk$set(echo = TRUE)

# see an overview of the data (glimpse is an alternative to str)
# str(df)
glimpse(df)
```

```{r view data}
knitr::opts_chunk$set(echo = TRUE)

# view the df in tabular form
view(df)
```


```{r data dimension}
knitr::opts_chunk$set(echo = TRUE)

# check the dimension of the data
dim(df)
```

```{r data summary}
knitr::opts_chunk$set(echo = TRUE)

# see a summary statistic of the data
summary(df)
```

```{r data null}
knitr::opts_chunk$set(echo = TRUE)

# check for null values in the data
colSums(is.na(df))
```

### Transform data
```{r new data instance}
knitr::opts_chunk$set(echo = TRUE)

# create a new instance of the data
df_clean = df
```

```{r datatype convert}
knitr::opts_chunk$set(echo = TRUE)

# convert columns to categorical according to the docs
# job, marital, education, default, housing, loan, contact, day, month, poutcome
df_clean <- df_clean %>%
  mutate_at(vars(job, marital, education, default, housing, loan, contact, day, month, poutcome), as.factor)
```

```{r data check}
knitr::opts_chunk$set(echo = TRUE)

# check if the operation was successful
glimpse(df_clean)
```

**Create new columns**
```{r create columns}
knitr::opts_chunk$set(echo = TRUE)

# create age ranges
df_clean$age_range <- cut(df_clean$age, breaks = c(0, 18, 30, 40, 50, 60, Inf), 
                        labels = c("0-18yrs", "19-30yrs", "31-40yrs", "41-50yrs", "51-60yrs", "60+yrs"))

# create call duration ranges
df_clean$call_duration <- cut(df_clean$duration, breaks = c(0, 10, 20, 30, 40, 50, 60, Inf), 
                                labels = c("0-10mins", "11-20mins", "21-30mins", "31-40mins", "41-50mins", "51-60mins", "1hr+"))


df_clean <- df_clean %>%
  mutate_at(vars(age_range, call_duration), as.factor)
```

### Analyze categorical variables
```{r cat coluns}
knitr::opts_chunk$set(echo = TRUE)

# select categorical columns
categorical_cols <- df_clean %>%
  select_if(is.factor)
```

```{r val counts}
knitr::opts_chunk$set(echo = TRUE)

# see the number of unique values in each column of the data
sapply(categorical_cols, function(x) if(is.factor(x)) nlevels(x) else length(unique(x)))
```

```{r age bar}
knitr::opts_chunk$set(echo = TRUE)

# Get frequency table of age range
age_table <- table(categorical_cols$age_range)

# Create bar plot
barplot(age_table, main = "Distribution of Categories", 
        xlab = "Age", ylab = "Frequency")
```

```{r call duration bar}
knitr::opts_chunk$set(echo = TRUE)

# Get frequency table of call_duration range
call_duration_table <- table(categorical_cols$call_duration)

# Create bar plot
barplot(call_duration_table, main = "Distribution of Categories", 
        xlab = "Call Duration", ylab = "Frequency")
```

```{r job bar}
knitr::opts_chunk$set(echo = TRUE)

# Get frequency table of job range
job_table <- table(categorical_cols$job)

# Create bar plot
barplot(job_table, main = "Distribution of Categories", 
        xlab = "Job Type", ylab = "Frequency", las = 2)
```

```{r marital bar}
knitr::opts_chunk$set(echo = TRUE)

# Get frequency table of marital range
marital_table <- table(categorical_cols$marital)

# Create bar plot
barplot(marital_table, main = "Distribution of Categories", 
        xlab = "Marital Status", ylab = "Frequency")
```

```{r education bar}
knitr::opts_chunk$set(echo = TRUE)

# Get frequency table of education range
education_table <- table(categorical_cols$education)

# Create bar plot
barplot(education_table, main = "Distribution of Categories", 
        xlab = "Education Status", ylab = "Frequency")
```

```{r default bar}
knitr::opts_chunk$set(echo = TRUE)

# Get frequency table of default range
default_table <- table(categorical_cols$default)

# Create bar plot
barplot(default_table, main = "Distribution of Categories", 
        xlab = "Default", ylab = "Frequency")
```

```{r housing bar}
knitr::opts_chunk$set(echo = TRUE)

# Get frequency table of housing range
housing_table <- table(categorical_cols$housing)

# Create bar plot
barplot(housing_table, main = "Distribution of Categories", 
        xlab = "Housing", ylab = "Frequency")
```

```{r loan bar}
knitr::opts_chunk$set(echo = TRUE)

# Get frequency table of loan range
loan_table <- table(categorical_cols$loan)

# Create bar plot
barplot(loan_table, main = "Distribution of Categories", 
        xlab = "Loan", ylab = "Frequency")
```

```{r contact bar}
knitr::opts_chunk$set(echo = TRUE)

# Get frequency table of contact range
contact_table <- table(categorical_cols$contact)

# Create bar plot
barplot(contact_table, main = "Distribution of Categories", 
        xlab = "Contact", ylab = "Frequency")
```

```{r day bar}
knitr::opts_chunk$set(echo = TRUE)

# Get frequency table of day range
day_table <- table(categorical_cols$day)

# Create bar plot
barplot(day_table, main = "Distribution of Categories", 
        xlab = "Day", ylab = "Frequency", las = 2)
```

```{r month bar}
knitr::opts_chunk$set(echo = TRUE)

# order month
month_order <- c("jan", "feb", "mar", "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec")
categorical_cols$month <- factor(categorical_cols$month, levels = month_order)

# Get frequency table of month range
month_table <- table(categorical_cols$month)

# Create bar plot
barplot(month_table, main = "Distribution of Categories", 
        xlab = "Month", ylab = "Frequency", 
        las = 2)
```

```{r poutcome bar}
knitr::opts_chunk$set(echo = TRUE)

# Get frequency table of poutcome range
poutcome_table <- table(categorical_cols$poutcome)

# Create bar plot
barplot(poutcome_table, main = "Distribution of Categories", 
        xlab = "Previous Outcome", ylab = "Frequency")
```

```{r all cat bar}
knitr::opts_chunk$set(echo = TRUE)

# Filter categorical columns
# categorical_plot <- names(categorical_cols)[sapply(categorical_cols, is.factor)]

# Create bar plots for each categorical column
# plot_list <- lapply(categorical_plot, function(col) {
#   plot_data <- as.data.frame(table(categorical_cols[, col]))
#   plot <- ggplot(plot_data, aes_string(x = names(plot_data)[1], y = names(plot_data)[2])) +
#     geom_bar(stat = "identity") +
#     ggtitle(col) +
#     xlab(col) +
#     ylab("Count") +
#     theme(axis.text.x = element_text(angle = 45, hjust = 1))
#   plot
# })

# grid.arrange(grobs = plot_list, ncol = 2)
```


### Analyze numerical variables
```{r num coluns}
knitr::opts_chunk$set(echo = TRUE)

# select numerical columns
numerical_cols <- df_clean %>%
  select_if(is.numeric)
```

```{r age hist}
knitr::opts_chunk$set(echo = TRUE)

ggplot(numerical_cols, aes(x = age)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "white") +
  labs(x = "Age Distribution", y = "Count") +
  ggtitle("Distribution of Age")
```

```{r balance hist}
knitr::opts_chunk$set(echo = TRUE)

ggplot(numerical_cols, aes(x = balance)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "white") +
  labs(x = "Age Distribution", y = "Count") +
  ggtitle("Distribution of Balance")
```

```{r duration hist}
knitr::opts_chunk$set(echo = TRUE)

ggplot(numerical_cols, aes(x = duration)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "white") +
  labs(x = "Call Duration Distribution", y = "Count") +
  ggtitle("Distribution of Duration")
```

```{r campaign hist}
knitr::opts_chunk$set(echo = TRUE)

ggplot(numerical_cols, aes(x = campaign)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "white") +
  labs(x = "Campaign Distribution", y = "Count") +
  ggtitle("Distribution of Campaign")
```

```{r pdays hist}
knitr::opts_chunk$set(echo = TRUE)

ggplot(numerical_cols, aes(x = pdays)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "white") +
  labs(x = "Previous Day Contact Distribution", y = "Count") +
  ggtitle("Distribution of Previous Day Contact") +
  scale_x_continuous(limits = c(0, 60))
```

```{r previous hist}
knitr::opts_chunk$set(echo = TRUE)

ggplot(numerical_cols, aes(x = previous)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "white") +
  labs(x = "Previous Campaigh Contact Distribution", y = "Count") +
  ggtitle("Distribution of Previous Campaigh Contact")
```

```{r all num bar}
knitr::opts_chunk$set(echo = TRUE)

# Filter numerical columns
# numerical_plot <- names(numerical_cols)[sapply(numerical_cols, is.numeric)]

# Create list of histogram plots for each numerical column
# plot_list <- lapply(numerical_plot, function(col) {
#   plot <- ggplot(numerical_cols, aes_string(x = col)) +
#     geom_histogram(binwidth = 1, fill = "blue", color = "white") +
#     ggtitle(col) +
#     xlab(col) +
#     ylab("Count")
#   plot
# })

# Arrange plots in a grid
# grid.arrange(grobs = plot_list, ncol = 2)
```

```{r correlation plot}
knitr::opts_chunk$set(echo = TRUE)

# create correlation for numeric variables

# Calculate correlation matrix
cor_matrix <- cor(numerical_cols)

# Create correlation plot
corrplot(cor_matrix, method = "color")
```


## Model
This section of the notebook experiments with different categorical models to predict the outcome of th variable

**Create data for model**
```{r select columns}
knitr::opts_chunk$set(echo = TRUE)

model_df <- df_clean[, c("job", "marital","education", "default", "balance", "housing", "loan", "contact", "campaign", "pdays", "previous", "poutcome", "age_range", "call_duration", "y")]

view(model_df)
```

```{r see proportion of the outcome class}
knitr::opts_chunk$set(echo = TRUE)

model_df <- na.omit(model_df)
prop.table(table(model_df$y))
```

```{r set seed}
knitr::opts_chunk$set(echo = TRUE)

set.seed(2020)
```

```{r creste test_train data}
knitr::opts_chunk$set(echo = TRUE)

y <- model_df$y
test_index <- createDataPartition(y, times = 1, p = 0.25, list = FALSE)
test_set <- model_df[test_index,]

train_set <- model_df[-test_index,]
train_set_numeric <- train_set %>% mutate(y = factor(y))
test_set_numeric <- test_set %>% mutate(y = factor(y))
```

### Decision Trees
```{r train decision tree using rpart}
knitr::opts_chunk$set(echo = TRUE)

train_rpart <- train(y ~ ., method = "rpart", tuneGrid = data.frame(cp = seq(0, 0.05, len = 25)), data=train_set_numeric)
confusionMatrix(predict(train_rpart, test_set_numeric), test_set_numeric$y)$overall["Accuracy"]
```

```{r decision confusion}
knitr::opts_chunk$set(echo = TRUE)

confusionMatrix(predict(train_rpart, test_set_numeric), test_set_numeric$y)
```

### Gradient Boosting Model
```{r train gbm}
knitr::opts_chunk$set(echo = TRUE)

# Define the tuning grid
train_gbm <- train(y ~ ., method = "gbm", 
                   tuneGrid = expand.grid(n.trees = c(50, 100, 150), 
                                          interaction.depth = c(1, 3, 5), 
                                          shrinkage = c(0.01, 0.1, 0.5),
                                          n.minobsinnode = c(5, 10, 20)), 
                   data = train_set_numeric)

confusionMatrix(predict(train_gbm, test_set_numeric), test_set_numeric$y)$overall["Accuracy"]
```

```{r see gbm confusion}
knitr::opts_chunk$set(echo = TRUE)

confusionMatrix(predict(train_gbm, test_set_numeric), test_set_numeric$y)
```

### Neural Network
```{r train neural network}
knitr::opts_chunk$set(echo = TRUE)

# Define the tuning grid
train_nnet <- train(y ~ ., method = "nnet",
                    tuneGrid = expand.grid(size = c(5, 10, 20),
                                           decay = c(0.1, 0.5, 1)),
                    data = train_set_numeric)

confusionMatrix(predict(train_nnet, test_set_numeric), test_set_numeric$y)$overall["Accuracy"]
```

```{r see nnet confusion}
knitr::opts_chunk$set(echo = TRUE)

confusionMatrix(predict(train_nnet, test_set_numeric), test_set_numeric$y)
```

## Test Data Prediction and Evaluation

**It is evident the best performing algorithm is the neural network algorithm according to the accuracy evaluation for all the algorithms used. We will proceed to import our test data, transform it and then evaluate the performance of the model against it.**
```{r test data}
knitr::opts_chunk$set(echo = TRUE)

df_final <- read.csv(file ='train.csv')
df_final_clean = df

# convert columns to categorical according to the docs
df_final_clean <- df_final_clean %>%
  mutate_at(vars(job, marital, education, default, housing, loan, contact, day, month, poutcome), as.factor)

# create age ranges
df_final_clean$age_range <- cut(df_final_clean$age, breaks = c(0, 18, 30, 40, 50, 60, Inf), 
                        labels = c("0-18yrs", "19-30yrs", "31-40yrs", "41-50yrs", "51-60yrs", "60+yrs"))

# create call duration ranges
df_final_clean$call_duration <- cut(df_final_clean$duration, breaks = c(0, 10, 20, 30, 40, 50, 60, Inf), 
                                labels = c("0-10mins", "11-20mins", "21-30mins", "31-40mins", "41-50mins", "51-60mins", "1hr+"))

# convert columns to categorical
df_final_clean <- df_final_clean %>%
  mutate_at(vars(age_range, call_duration), as.factor)

# pick model data
final_model_df <- df_final_clean[, c("job", "marital","education", "default", "balance", "housing", "loan", "contact", "campaign", "pdays", "previous", "poutcome", "age_range", "call_duration", "y")]

# create model data
final_model_df <- na.omit(final_model_df)
y <- final_model_df$y
final_test_index <- createDataPartition(y, times = 1, p = 1, list = FALSE)

final_test_set <- model_df[final_test_index,]
pred_data <- final_test_set %>% mutate(y = factor(y))

# predict and evaluate model
predictions <- predict(train_nnet, newdata = pred_data)
confusionMatrix(predictions, reference = pred_data$y)
```